Predictive Modeling of I/O Characteristics \\ in High Performance Computing Systems

### 2018-01-08 Paper_[2017-11]_HPC/paper.tex
Finished submitted draft of ACM SE paper.

\vspace{-10pt}
\subsubsection{Delaunay}
\vspace{-10pt}
The Delaunay method of interpolation is a well studied geometric
technique for producing an interpolant \shortcite{lee1980two}. The
Delaunay triangulation of a set of data points into simplices is such
that the sphere defined by the vertices of each simplex contains no
data points in the sphere's interior. For a $d$-simplex S with
vertices $v^{(0)}, v^{(1)}, \ldots, v^{(d)}$, $x \in S$, and data
values $f(v^{(i)})$, $i=0$,$\ldots$,$d$, $x$ is a unique convex
combination of the vertices,

@@

\subsubsection{Multidimensional Analysis}
\vspace{-10pt}
This procedure combines random selection of training and testing
splits with changes in the input dimension and the ratio of training
size to testing size. Given an input data matrix with $n$ rows
(points) and $d$ columns (components), MDA proceeds as follows:
\begin{enumerate}
\item For all $k = 1$, $\ldots$, $d$ and for all nonempty subsets $F
  \subset \{1, 2, \ldots, d\}$, reduce the input data to points $(z,
  f_F(z))$ with $z \in \mathbb{R}^k$ and $f_F(z) = E[ \{ f(x^{(i)})
    \mid (x^{(i)}_F = z) \} ]$, where $E[\cdot]$ denotes the mean and
  $x^{(i)}_F$ is the subvector of $x^{(i)}$ indexed by $F$.
\item For all $r$ in $\{5, 10, \ldots, 95\}$, generate $N$ random
  splits $(train, test)$ of the reduced data with $r$ percentage for
  training and $100 - r$ percentage for testing.
\item When generating each of $N$ random $(train, test)$ splits,
  ensure that all points from $test$ are in the convex hull of points
  in $train$ (to prevent extrapolation); also ensure that the points
  in $train$ are well spaced.
\end{enumerate}

In order to ensure that the testing points are in the convex hull of
the training points, the convex hull vertices of each set of (reduced
dimension) points are forcibly placed in the training set. In order to
ensure that training points are well spaced, a statistical method for
picking points from \shortciteN{amos2014algorithm} is used:
\begin{enumerate}
\item Generate a sequence of all pairs of points
  ${(z^{(i_1)},z^{(j_1)}), (z^{(i_2)},z^{(j_2)}), \ldots}$ sorted by
  ascending pairwise Euclidean distance between points, so that
  $||z^{(i_k)}-z^{(j_k)}||_2 \leq ||z^{(i_{k+1})}-z^{(j_{k+1})}||_2$.
\item Sequentially remove points from candidacy until only $|train|$
  remain by randomly selecting one point from the pair $(z^{(i_m)},
  z^{(j_m)})$ for $m = 1,\ldots$ if both $z^{(i_m)}$ and $z^{(j_m)}$
  are still candidates for removal.
\end{enumerate}

Given the large number of constraints, level of reduction, and use of
randomness in the MDA procedure, occasionally $N$ unique training /
testing splits may not be created or may not exist. In these cases, if
there are fewer than $N$ possible splits, then deterministically
generated splits are used. Otherwise after $3N$ attempts, only the
unique splits are kept for analysis. The MDA procedure has been
implemented in Python\#3 while most regression and interpolation
methods are Fortran wrapped with Python. All randomness has been
seeded for repeatability.

@@

A na\"{\i}ve multivariate prediction technique such as nearest
neighbor could experience relative errors in the range $\displaystyle
[0, \big(\max_x f(x) - \min_x f(x)\big) / \min_x f(x) ]$ when modeling
a nonnegative function $f(x)$ from data. The IOzone mean data response
values span three orders of magnitude (as can be seen in Table
\ref{tab:data_type}) while variance data response values span six
orders of magnitude. It is expected therefore, that all studied
multivariate models perform better than a na\"{\i}ve approach,
achieving relative errors strictly less than $10^3$ for throughput
mean and $10^6$ for throughput sample variance. Ideally, models will
yield relative errors significantly smaller than 1.

@@

The present results demonstrate that a straightforward application of
multivariate modeling techniques can be used to effectively predict
HPC system performance. Some modeling effort on the part of a systems
engineer combined with a significant amount of experimentation (weeks
of CPU time for the IOzone data used here) can yield a model capable
of accurately tuning an HPC system to the desired performance
specification, although qualitatively correct predictions can be
achieved with mush less (10\%, say) effort.

>>>>

\vspace{-10pt}
\subsubsection{Delaunay}
\vspace{-10pt}
The Delaunay method of interpolation is a well studied geometric
technique for producing an interpolant \shortcite{lee1980two}. The
Delaunay triangulation of a set of data points into simplices is such
that the sphere defined by the vertices of each simplex contains no
data points in the sphere's interior. For a $d$-simplex S with
vertices $v^{(0)}$, $v^{(1)}$, $\ldots$, $v^{(d)}$, $x \in S$, and
data values $f(v^{(i)})$, $i=0$, $\ldots$, $d$, $x$ is a unique convex
combination of the vertices,

@@

\subsubsection{Multidimensional Analysis}
\vspace{-10pt}
This procedure combines random selection of training and testing
splits with changes in the input dimension and the ratio of training
size to testing size. Given an input data matrix with $n$ rows
(points) and $d$ columns (components), MDA proceeds as follows:
\begin{enumerate}
\item For all $k = 1$, $\ldots$, $d$ and for all nonempty subsets $F
  \subset \{1, 2, \ldots, d\}$, reduce the input data to points $(z,
  f_F(z))$ with $z \in \mathbb{R}^k$ and $f_F(z) = E\bigl[ \bigl\{
    f\bigl(x^{(i)}\bigr) \bigm| \bigl(x^{(i)}_F = z\bigr) \bigr\}
    \bigr]$, where $E[\cdot]$ denotes the mean and $x^{(i)}_F$ is the
  subvector of $x^{(i)}$ indexed by $F$.
\item For all $r$ in $\{5, 10, \ldots, 95\}$, generate $N$ random
  splits $(train, test)$ of the reduced data with $r$ percentage for
  training and $100 - r$ percentage for testing.
\item When generating each of $N$ random $(train, test)$ splits,
  ensure that all points from $test$ are in the convex hull of points
  in $train$ (to prevent extrapolation); also ensure that the points
  in $train$ are well spaced.
\end{enumerate}

In order to ensure that the testing points are in the convex hull of
the training points, the convex hull vertices of each set of (reduced
dimension) points are forcibly placed in the training set. In order to
ensure that training points are well spaced, a statistical method for
picking points from \shortciteN{amos2014algorithm} is used:
\begin{enumerate}
\item Generate a sequence of all pairs of points
  $\bigl(z^{(i_1)},z^{(j_1)}\bigr), \bigl(z^{(i_2)},z^{(j_2)}\bigr),
  \ldots$ sorted by ascending pairwise Euclidean distance between
  points, so that $\bigl|\bigl|z^{(i_k)}-z^{(j_k)}\bigr|\bigr|_2 \leq
  \bigl|\bigl|z^{(i_{k+1})}-z^{(j_{k+1})}\bigr|\bigr|_2$.
\item Sequentially remove points from candidacy until only $|train|$
  remain by randomly selecting one point from the pair
  $\bigl(z^{(i_m)}, z^{(j_m)}\bigr)$ for $m = 1,\ldots$ if both
  $z^{(i_m)}$ and $z^{(j_m)}$ are still candidates for removal.
\end{enumerate}

Given the large number of constraints, level of reduction, and use of
randomness in the MDA procedure, occasionally $N$ unique
training/testing splits may not be created or may not exist. In these
cases, if there are fewer than $N$ possible splits, then
deterministically generated splits are used. Otherwise after $3N$
attempts, only the unique splits are kept for analysis. The MDA
procedure has been implemented in Python\#3 while most regression and
interpolation methods are Fortran wrapped with Python. All randomness
has been seeded for repeatability.

@@

A na\"{\i}ve multivariate prediction technique such as nearest
neighbor could experience relative errors in the range $\displaystyle
[0, \big(\max_x f(x) - \min_x f(x)\big) / \min_x f(x) ]$ when modeling
a nonnegative function $f(x)$ from data. The IOzone mean data response
values span three orders of magnitude (as can be seen in Table
\ref{tab:data_type}) while variance data response values span six
orders of magnitude. It is expected therefore, that all studied
multivariate models perform better than a na\"{\i}ve approach,
achieving relative errors strictly less than $10^3$ for throughput
mean and $10^6$ for throughput variance. Ideally, models will
yield relative errors significantly smaller than 1.

@@

The present results demonstrate that a straightforward application of
multivariate modeling techniques can be used to effectively predict
HPC system performance. Some modeling effort on the part of a systems
engineer combined with a significant amount of experimentation (days
of CPU time for the IOzone data used here) can yield a model capable
of accurately tuning an HPC system to the desired performance
specification, although qualitatively correct predictions can be
achieved with mush less (10\%, say) effort.
